---
title: "Assignment B"
format: 
   html:
     embed-resources: true
editor: visual
---

## Exercises 1. Vectors

1.  Create the vectors:

    (a) $(1, 2, 3, \dots, 19, 20)$

    (b) $(20, 19, \dots, 2, 1)$

    (c) $(1, 2, 3, \dots, 19, 20, 19, 18, \dots, 2, 1)$

    (d) $(4, 6, 3)$ and assign it to the name **tmp**. <br> For parts (e), (f) and (g) look at the help for the function rep.

    (e) $(4, 6, 3, 4, 6, 3, \dots, 4, 6, 3)$ where there are 10 occurrences of 4.

    (f) $(4, 6, 3, 4, 6, 3, \dots, 4, 6, 3, 4)$ <br>where there are 11 occurrences of 4, 10 occurrences of 6 and 10 occurrences of 3.

    (g) $(4, 4, \dots, 4, 6, 6, \dots, 6, 3, 3, \dots,3)$ <br> where there are 10 occurrences of 4, 20 occurrences of 6 and 30 occurrences of 3.

```{r}
#|
```

```{r}
#| label: 1(a-g)
#| echo: true

#a
p_1a <- 1:20
p_1a
#b
p_1b <- 20:1
p_1b
#c
p_1c <- c(1:20,19:1)
p_1b

#d & e
#############################
# from chatgpt
# answer 1
tmp <- c(4, 6, 3)
v <- rep(tmp, length.out = 10 * length(tmp))  # Repeat enough times
## checkit
v
length(which(v==4))

## chatGPT answer 2
v <- v[1:(which(cumsum(v == 4) == 10)[1] * length(tmp)/sum(tmp == 4))]  

# Trim to exactly 10 occurrences of 4
## Check it
v
length(which(v==4))
##############################
## my answer

p_1e <- rep(tmp,10)
p_1e
length(which(p_1e==4))

############################

#f

p_1f <- rep(tmp, l=31)
p_1f
length(which(p_1f==4))
length(which(p_1f==6))
length(which(p_1f==3))

#g
p_1g <- rep(tmp, times=c(10,20,30))
p_1g
length(which(p_1g==4))
length(which(p_1g==6))
length(which(p_1g==3))



```

2.  Create a vector of the values of $e^x cos(x)$ at $x = 3, 3.1, 3.2, \dots, 6.$

```{r}
#| label: prob 2
library(base)
library(graphics)

tmp <- seq(3, 6, by=0.1)
p_2 <- exp(tmp)*cos(tmp)

plot(tmp,exp(tmp), type = "l", main = "exp(tmp)")
plot(tmp,cos(tmp), type = "l")
plot(tmp,p_2,type = "l")


```

3.  Create the following vectors:

    (a) $(0.1^30.2^1, 0.1^6 0.2^4, \dots, 0.1^{36}0.2^{34})$

    (b) $(2, \frac{2^2}{2}, \frac{2^3}{3}, \dots, \frac{2^{25}}{25})$

```{r}
#| label: Prob 3a
#| 

options(scipen = 0)

p_3a <- (0.1^seq(3,36,by=3))*(0.2^seq(1,34,by=3))
p_3a

a <- (0.1^seq(3,36,by=3))
a
b <- (0.2^seq(1,34,by=3))
b

a*b == p_3a

```

4.  Calculate the following:

    (a) $\displaystyle \sum _{i=10}^{100} (i^3 + 4i^2)$

    (b) $\displaystyle \sum_{i=1}^{25} \left ( \frac{2^i}{i} + \frac{3^i}{i^2} \right )$

<p style="page-break-before:always">

</p>

```{r}
#| label: Prob 4a

tmp <- 10:100
tmp^3 + 4*tmp^2
plot(tmp, tmp^3 + 4*tmp^2, pch=20 ,main = "prob 4a")
sum(tmp^3+4*tmp^2)

```

5.  Use the function paste to create the following character vectors of length 30:

    (a) ("label 1", "label 2", ... , "label 30")<br> Note that there is a single space between **label** and the number following.

    (b) ("fn1", "fn2", ... , "fn30"). <br> In this case, there is no space between **fn** and the number following.

6.  Execute the following lines which create two vectors of random integers which are chosen with replacement from the integers $0, 1, \dots , 999$. Both vectors have length 250.

::: {style="margin-left:2.5cm"}
set.seed(50)

xVec \<- sample(0:999, 250, replace=T)\
yVec \<- sample(0:999, 250, replace=T)

Suppose x = (x_1, x_2, \dots, x_n) denotes the vector xVec and\
y = (y_1, y_2, \dots, y_n) denotes the vector yVec.
:::

::: {style="margin-left:1.5cm"}
(a) Create the vector $(y_2 - x_1, \dots,  y_n - x_{n−1})$.

(b) Create the vector $(\frac{sin(y_1)}{cos(x_2)}, \frac{sin(y_2)}{cos(x_3)}, \dots, \frac{sin(y_{n−1})}{cos(x_n)})$

(c) Create the vector $(x_1 + 2x_2 - x_3, x_2 + 2x_3 - x_4, \dots, x_{n−2} + 2x_{n−1} - x_n)$

(d) Calculate $\displaystyle \sum_{i=1}^{n-1} \frac{e^{-x_{i+1}}}{x_i + 10}$
:::

```{r}
#| label: prob 6

## setup

xVec <- sample(0:999, 250, replace=T)
yVec <- sample(0:999, 250, replace=T)

## for illustration

xv1 <- xVec[1:25]
yv1 <- yVec[1:25]

###a
yv1
yv1[-1]

```

7.  This question uses the vectors xVec and yVec created in the previous question and the functions sort(), order(), mean(), sqrt(), sum(), and abs().

    (a) Pick out the values in yVec which are \> 600.

    (b) What are the index positions in yVec of the values which are \> 600?

    (c) What are the values in xVec which correspond to the values in yVec which are \> 600?<br> (By correspond, we mean at the same index positions.)

    (d) Create the vector $(|x_1 - \bar{x}|^{1/2}, \;\;|x_2 - \bar{x}|^{1/2},\;\; \dots\;,\;\; |x_n - \bar{x}|^{1/2} )$<br> where $\bar{x}$ denotes the mean of the vector $x = (x_1, x_2, \dots, x_n)$.

    (e) How many values in yVec are within 200 of the maximum value of the terms in yVec?

    (f) How many numbers in xVec are divisible by 2?<br> (Note that the modulo operator is denoted **%%**.)

    (g) Sort the numbers in the vector xVec in the order of increasing values in yVec.

    (h) Pick out the elements in yVec at index positions 1, 4, 7, 10, 13, ...

8.  By using the function cumprod() or otherwise, calculate

    ::: {style="margin-left: 2cm"}
    $1 + \frac{2}{3} + \left (\frac{2}{3}\;\frac{4}{5} \right ) + \left (\frac{2}{3}\;\frac{4}{5}\;\frac{6}{7} \right ) + \dots +$ $\left (\frac{2}{3}\;\frac{4}{5}\;\dots \frac{38}{39} \right )$
    :::

```{r}
#| prob_8

1+sum(cumprod(seq(2,38,b=2)/seq(3,39,b=2)))

seq(2,38,b=2)

seq(3,39,b=2)

seq(2,38,b=2)/seq(3,39,b=2)

cumprod(seq(2,38,b=2)/seq(3,39,b=2))

sum(cumprod(seq(2,38,b=2)/seq(3,39,b=2)))

```
